Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-26 Manuscript/PNAS-Adams-Submitted.Rmd

---
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry of species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition the data by the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @MartinsHansen1997; @OMeara_et_al2006;  @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. PCMs are predicated on the notion that phylogenetic signal -- the tendancy for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covaration among taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008].

@@

The appeal of Pagel's $\lambda$ and Blomberg's *K* as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or *K*, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether such comparisons are analytically appropriate, and whether these statistics are, or can be, converted to effect sizes for comparative analyses across datasets. To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ is a random variable that follows the distribution of a Bernoulli probability parameter [@Forbes2011]; i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ will be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 [@Forbes2011]. Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging.

By contrast, for Blomberg's *K*, which is positively unbounded, we might expect that for any $\lambda$ used to generate data, estimates of *K* might be a random variable that follows a normal distribution, with values distributed symmetrically about the input value [@Forbes2011]. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's *K* might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's *K*, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the reliability of these statistics as effect sizes.  Furthermore, the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages.

>>>>

---
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry of species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition the data by the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @MartinsHansen1997; @OMeara_et_al2006;  @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. PCMs are predicated on the notion that phylogenetic signal -- the tendency for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covariation among taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008].

@@

The appeal of Pagel's $\lambda$ and Blomberg's *K* as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or *K*, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether such comparisons are analytically appropriate, and whether these statistics are, or can be, converted to effect sizes for comparative analyses across datasets. To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ is a random variable that follows the distribution of a Bernoulli probability parameter [@Forbes2011]; i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ would be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 [@Forbes2011]. Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging.

By contrast, for Blomberg's *K*, which is positively unbounded, we might expect that for any $\lambda$ used to generate data, estimates of *K* might be a random variable that follows a normal distribution, with values distributed symmetrically [@Forbes2011]. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's *K* might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's *K*, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the reliability of these statistics as effect sizes.  Furthermore, the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages.

### 2020-08-26 Manuscript/PNAS-Adams-Submitted.tex

Investigating macroevolutionary patterns of trait variation requires a
phylogenetic perspective, because the shared ancestry of species
violates the assumption of independence among trait values that is
common for statistical tests (1, 2). Accounting for this evolutionary
non-independence is the purview of \emph{phylogenetic comparative
methods} (PCMs): a suite of analytical tools that condition the data by
the phylogenetic relatedness of observations (3--10). PCMs are
predicated on the notion that phylogenetic signal -- the tendancy for
closely related species to display similar trait values -- is present in
cross-species datasets (1, 11, 12). Indeed, under numerous evolutionary
models, phylogenetic signal is expected, as stochastic character change
along the hierarchical structure of the tree of life generates trait
covaration among taxa (1, 12, 13).

@@

The appeal of Pagel's \(\lambda\) and Blomberg's \emph{K} as descriptive
statistics is that they provide a basis for interpreting ``weak'' versus
``strong'' phylogenetic signal; i.e., small versus large values of
\(\hat{\lambda}\) or \emph{K}, respectively, in a comparative sense
(28--30). Nonetheless, an important question that has yet to be
considered is whether such comparisons are analytically appropriate, and
whether these statistics are, or can be, converted to effect sizes for
comparative analyses across datasets. To be statistics representing
phylogenetic signal, they should have reliable distributional
properties, which could be revealed with simulation experiments. For
instance, as a proportional random variable bounded by \(0\) and \(1\),
we might expect that \(\hat{\lambda}\) is a random variable that follows
the distribution of a Bernoulli probability parameter (31); i.e., branch
lengths in a tree are scaled proportionally to the probability that data
arise from a BM process. Given a known \(\lambda\) value used to
generate random data on a tree, we would also expect that the mean of an
empirical sampling distribution of \(\hat{\lambda}\) would approximately
equal \(\lambda\); the dispersion of \(\hat{\lambda}\) would be largest
at intermediate values of \(\lambda\), \(\hat{\lambda}\) would be
predictable over the range of \(\lambda\) with respect to tree size; the
distribution of \(\hat{\lambda}\) would be symmetric at intermediate
values of \(\lambda\) and more skewed toward values of 0 or 1; and that
the distribution of \(\hat{\lambda}\) will be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (31). Prior work (18) seems to support some of these conjectures,
based superficially on statistical moments for a given tree size (mean,
variance, skewness, and kurtosis; see Fig. 2 of ref. (18)). However,
because the ``strength of Brownian motion'' was simulated as a varied
weighted-average of data simulated on trees with \(\lambda=0\) and
\(\lambda=1\) and not as prescribed values of \(\lambda\) (18),
interpretation of these patterns is challenging.

By contrast, for Blomberg's \emph{K}, which is positively unbounded, we
might expect that for any \(\lambda\) used to generate data, estimates
of \emph{K} might be a random variable that follows a normal
distribution, with values distributed symmetrically about the input
value (31). This attribute seemed less reasonable based on the
simulations performed by Münkemüller et al.~(18), which suggested that
distributions were positively skewed and that Blomberg's \emph{K} might
not behave as a statistic that follows a normal distribution. However,
because their simulations used a weighted combination of simulated
phylogenetic signal strengths, strong inferences are not possible (and
distributional attributes were not the intended result of their
simulations). Thus, for both Pagel's \(\lambda\) or Blomberg's \emph{K},
evaluation of statistical moments across a range of \(\lambda\) used to
generate data would be valuable for adjudicating the reliability of
these statistics as effect sizes. Furthermore, the expected values of
these statistics appear to vary with tree size (18), making comparisons
across studies challenging. Therefore, transformation of these
statistics into \(Z\)-scores would allow evaluation of the efficacy of
each statistic to yield effect sizes that could be used for comparisons
of the strength of phylogenetic signal across traits and lineages.

>>>>

Investigating macroevolutionary patterns of trait variation requires a
phylogenetic perspective, because the shared ancestry of species
violates the assumption of independence among trait values that is
common for statistical tests (1, 2). Accounting for this evolutionary
non-independence is the purview of \emph{phylogenetic comparative
methods} (PCMs): a suite of analytical tools that condition the data by
the phylogenetic relatedness of observations (3--10). PCMs are
predicated on the notion that phylogenetic signal -- the tendency for
closely related species to display similar trait values -- is present in
cross-species datasets (1, 11, 12). Indeed, under numerous evolutionary
models, phylogenetic signal is expected, as stochastic character change
along the hierarchical structure of the tree of life generates trait
covariation among taxa (1, 12, 13).

@@

The appeal of Pagel's \(\lambda\) and Blomberg's \emph{K} as descriptive
statistics is that they provide a basis for interpreting ``weak'' versus
``strong'' phylogenetic signal; i.e., small versus large values of
\(\hat{\lambda}\) or \emph{K}, respectively, in a comparative sense
(28--30). Nonetheless, an important question that has yet to be
considered is whether such comparisons are analytically appropriate, and
whether these statistics are, or can be, converted to effect sizes for
comparative analyses across datasets. To be statistics representing
phylogenetic signal, they should have reliable distributional
properties, which could be revealed with simulation experiments. For
instance, as a proportional random variable bounded by \(0\) and \(1\),
we might expect that \(\hat{\lambda}\) is a random variable that follows
the distribution of a Bernoulli probability parameter (31); i.e., branch
lengths in a tree are scaled proportionally to the probability that data
arise from a BM process. Given a known \(\lambda\) value used to
generate random data on a tree, we would also expect that the mean of an
empirical sampling distribution of \(\hat{\lambda}\) would approximately
equal \(\lambda\); the dispersion of \(\hat{\lambda}\) would be largest
at intermediate values of \(\lambda\), \(\hat{\lambda}\) would be
predictable over the range of \(\lambda\) with respect to tree size; the
distribution of \(\hat{\lambda}\) would be symmetric at intermediate
values of \(\lambda\) and more skewed toward values of 0 or 1; and that
the distribution of \(\hat{\lambda}\) would be more platykurtic at
intermediate values of \(\lambda\), becoming more leptokurtic toward 0
and 1 (31). Prior work (18) seems to support some of these conjectures,
based superficially on statistical moments for a given tree size (mean,
variance, skewness, and kurtosis; see Fig. 2 of ref. (18)). However,
because the ``strength of Brownian motion'' was simulated as a varied
weighted-average of data simulated on trees with \(\lambda=0\) and
\(\lambda=1\) and not as prescribed values of \(\lambda\) (18),
interpretation of these patterns is challenging.

By contrast, for Blomberg's \emph{K}, which is positively unbounded, we
might expect that for any \(\lambda\) used to generate data, estimates
of \emph{K} might be a random variable that follows a normal
distribution, with values distributed symmetrically (31). This attribute
seemed less reasonable based on the simulations performed by Münkemüller
et al.~(18), which suggested that distributions were positively skewed
and that Blomberg's \emph{K} might not behave as a statistic that
follows a normal distribution. However, because their simulations used a
weighted combination of simulated phylogenetic signal strengths, strong
inferences are not possible (and distributional attributes were not the
intended result of their simulations). Thus, for both Pagel's
\(\lambda\) or Blomberg's \emph{K}, evaluation of statistical moments
across a range of \(\lambda\) used to generate data would be valuable
for adjudicating the reliability of these statistics as effect sizes.
Furthermore, the expected values of these statistics appear to vary with
tree size (18), making comparisons across studies challenging.
Therefore, transformation of these statistics into \(Z\)-scores would
allow evaluation of the efficacy of each statistic to yield effect sizes
that could be used for comparisons of the strength of phylogenetic
signal across traits and lineages.

### 2020-08-26 Manuscript/SupportingInfo.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal},
  pdfauthor={Dean C. Adams, Erica K. Baken, and Michael L. Collyer},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Supporting Information for: A Standardized Effect Size for Evaluating
and Comparing the Strength of Phylogenetic Signal}
\author{Dean C. Adams, Erica K. Baken, and Michael L. Collyer}
\date{}

\begin{document}
\maketitle

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\includegraphics[width=0.95\linewidth]{SI-Title} \hfill\break
\hfill\break

\hypertarget{supplementary-information-for}{%
\subsection{Supplementary Information
for}\label{supplementary-information-for}}

\hfill\break

A Standardized Effect Size for Evaluating and Comparing the Strength of
Phylogenetic Signal \hfill\break \hfill\break Dean C. Adams, Erica K.
Baken, and Michael L. Collyer \hfill\break

\hfill\break

Dean C. Adams\hfill\break Email:
\href{mailto:dcadams@iastate.edu}{\nolinkurl{dcadams@iastate.edu}}\hfill\break

\hypertarget{this-pdf-file-includes}{%
\subsection{This PDF file includes:}\label{this-pdf-file-includes}}

\hfill\break

\begin{itemize}
\tightlist
\item
  Supplementary text
\item
  Figures S1 to S20
\item
  SI References
\end{itemize}
